---
title: "STA302 Report"
output:
  pdf_document: default
  html_notebook: default
---

# Group 173: Oscar Heath, Evelyn Hughes, Amy Ye

### Importing Data from NHANES package

```{r}
#import libraries
library(NHANES)
library(MASS)
library(car)
data(NHANES)

# select columns we want from NHANES
data <- subset(as.data.frame(NHANES), 
              select = c(BPSysAve, Gender, Age, BMI, 
                         HHIncomeMid, Poverty, Weight, Pulse, TotChol, PhysActiveDays, Race1, PhysActive,TVHrsDay))
data = na.omit(data)
set.seed(42)
s <- sample(1:nrow(data), 1000, replace=F)
train <- data[s, ]
test <- data[-s, ]
```

### Plot the data to see distributions:
```{r, echo=F, fig.width=8, fig.height=11}
num_cols <- ncol(data)
par(mfrow = c(ceiling(num_cols / 4), 4)) 

# loop through each column and plot the distribution
for (col_name in names(data)) {
  data_col <- data[[col_name]]
  
  # if the variable is numeric, then we will make a histogram
  if (is.numeric(data_col)) {
      hist(data_col, 
         main = paste("Histogram of", col_name), 
         xlab = col_name,
         col = "lightblue",
         border = "black")
         
  } else {
  # if categorical, make a barplot
  barplot(table(data_col), 
          main = paste("Bar Plot of", col_name), 
          xlab = col_name,
          ylab = "Count",
          col = "lightgreen",
          las = 2)
  }
}
par(mfrow = c(1, 1))
```

### Checking correlation between predictors:

```{r}
plot(data)
```

### Summary of BPSysAve, our response variable

```{r}
summary(data$BPSysAve)
```

### Graphing interaction terms:

```{r}
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)

# Make a scatterplot of blood pressure and age, coloured by gender
plot(data$Age, data$BPSysAve, 
     main = "Interaction of Age and Gender on Blood Pressure",
     xlab = "Age",
     ylab = "Systolic Blood Pressure",
     pch = 19,
     col = ifelse(data$Gender == "male", "blue", "red"))
# Add female line of best fit
female_fit <- lm(BPSysAve ~ Age, data = subset(data, Gender == "female"))
abline(female_fit, col = "red", lwd = 2)
# Add male line of best fit
male_fit <- lm(BPSysAve ~ Age, data = subset(data, Gender == "male"))
abline(male_fit, col = "blue", lwd = 2)

legend("topleft", 
       legend = c("Female", "Male"), 
       col = c("red", "blue"), 
       pch = 19, lwd = 2, bty = "n")

# Make a scatterplot of blood pressure and BMI
plot(data$BMI, data$BPSysAve, 
     main = "Interaction of BMI and Gender on Blood Pressure",
     xlab = "BMI",
     ylab = "Systolic Blood Pressure",
     pch = 19, 
     col = ifelse(data$Gender == "male", "blue", "red")) 
# Add female line of best fit
female_fit_bmi <- lm(BPSysAve ~ BMI, data = subset(data, Gender == "female"))
abline(female_fit_bmi, col = "red", lwd = 2)
# Add male line of best fit
male_fit_bmi <- lm(BPSysAve ~ BMI, data = subset(data, Gender == "male"))
abline(male_fit_bmi, col = "blue", lwd = 2)

legend("topleft", 
       legend = c("Female", "Male"), 
       col = c("red", "blue"), 
       pch = 19, lwd = 2, bty = "n")

# Split TotChol into a high and low category using the median, we will use this to visualize different slopes
median_chol <- median(data$TotChol)

# Make a scatterplot of blood pressure and cholestrol
plot(data$Age, data$BPSysAve, 
     main = "Interaction of Age and Total Cholesterol on Blood Pressure",
     xlab = "Age",
     ylab = "Systolic Blood Pressure",
     pch = 19, 
     col = ifelse(data$TotChol < median_chol, "red", "blue")) #colour based on group
# Add low cholestrol line of best fit
low_chol_fit <- lm(BPSysAve ~ Age, data = subset(data, TotChol < median_chol))
abline(low_chol_fit, col = "red", lwd = 2)
# Add high cholestrol line of best fit
high_chol_fit <- lm(BPSysAve ~ Age, data = subset(data, TotChol >= median_chol))
abline(high_chol_fit, col = "blue", lwd = 2)

legend("topleft", 
       legend = c("Low Cholesterol", "High Cholesterol"), 
       col = c("red", "blue"), 
       pch = 19, lwd = 2, bty = "n")

par(mfrow = c(1, 1))
```

### Fitting our preliminary model

```{r}
# Fit a model with our predictors, as well as some interesting interaction terms
model = lm(BPSysAve ~. + Gender:Age + Gender:BMI + Age:TotChol, data=data)
summary(model)
```

### Residual Plots:

```{r, echo=F, fig.width=8, fig.height=11}
# Get residuals and fitted values
e_hat <- resid(model)
y_hat <- fitted(model)

# Plot residuals
par(mfrow=c(4,3))
plot(data$BPSysAve ~ y_hat, ylab="True Systolic Blood Pressure", xlab="Fitted")
plot(e_hat ~ y_hat, ylab="Residuals", xlab="Fitted")
for(i in c(2,3,4,5,6,7,8,9,10,11,12)){
  plot(e_hat ~ data[,i], ylab="Residuals", xlab=names(data)[i])
}
plot(e_hat, ylab="Residuals", xlab="Observation number")
qqnorm(e_hat)
qqline(e_hat)
```
```{r}
vif(model, type="predictor")
```
### VIF seems high for some variables: performing a Partial F Test to see if we can drop Weight and HHIncomeMid
```{r}
full_model = model
reduced_model = lm(BPSysAve ~ Gender + Age + BMI + Poverty + Pulse + TotChol + PhysActiveDays + Race1 + PhysActive + TVHrsDay + Gender:Age + Gender:BMI + Age:TotChol, data=data)
anova(reduced_model, full_model)
vif(reduced_model, type="predictor")
```
### Test indicates reduced_model is nearly as good as full model, VIFs are now all < 2

### Let's find some outliers we can possible prune
```{r}
#outlier pruning
n <- nrow(data)
p <- length(coef(reduced_model)) - 1

# leverage cutoff
hcut <- 2*((p+1)/n)

# cook's distance cutoff
cookcut <- qf(0.5, p+1, n-p-1)

h_ii <- hatvalues(reduced_model)

# standardized residuals
r_i <- rstandard(reduced_model)

# cook's distance 
D_i <- cooks.distance(reduced_model)

# get residuals
residuals_i <- residuals(reduced_model)

# set a cutoff based on the standardized residuals
negative_residual_cutoff <- -2

# find large negative outliers
treated_outlier_indices <- which(r_i < negative_residual_cutoff)

leverage_outliers <- which(h_ii > hcut)
cooks_outliers <- which(D_i > cookcut)

statistical_influencers <- intersect(leverage_outliers, cooks_outliers)
treated_outliers <- which(r_i < negative_residual_cutoff)

# get all outliers
all_outlier_indices <- unique(union(treated_outliers, statistical_influencers))

cat("--- Final Indices to Consider for Removal ---\n")
print(all_outlier_indices)
```

```{r}
data_cleaned <- data[-all_outlier_indices, ]
cat("Original sample size (n):", nrow(data), "\n")
cat("Number of observations dropped:", length(all_outlier_indices), "\n")
cat("Cleaned sample size (n_cleaned):", nrow(data_cleaned), "\n")
```

```{r}
model_formula <- BPSysAve ~ Gender + Age + BMI + Poverty + Pulse + TotChol + PhysActiveDays + Race1 + PhysActive + TVHrsDay + Gender:Age + Gender:BMI + Age:TotChol
reduced_model_cleaned <- lm(model_formula, data = data_cleaned)
summary(reduced_model_cleaned)
```

```{r, echo=F, fig.width=8, fig.height=11}
# Get residuals and fitted values
e_hat <- resid(reduced_model_cleaned)
y_hat <- fitted(reduced_model_cleaned)

# Plot residuals
par(mfrow=c(4,3))
plot(data_cleaned$BPSysAve ~ y_hat, ylab="True Systolic Blood Pressure", xlab="Fitted")
plot(e_hat ~ y_hat, ylab="Residuals", xlab="Fitted")
for(i in c(2,3,4,5,6,7,8,9,10,11,12)){
  plot(e_hat ~ data_cleaned[,i], ylab="Residuals", xlab=names(data_cleaned)[i])
}
plot(e_hat, ylab="Residuals", xlab="Observation number")
qqnorm(e_hat)
qqline(e_hat)
```
```{r}
logBP = log(data_cleaned$BPSysAve)
reduced_model_cleaned_logged <- lm(logBP ~ Gender + Age + BMI + Poverty + Pulse + TotChol + PhysActiveDays + Race1 + PhysActive + TVHrsDay + Gender:Age + Gender:BMI + Age:TotChol, data = data_cleaned)
summary(reduced_model_cleaned_logged)
```


```{r, echo=F, fig.width=8, fig.height=11}
# Get residuals and fitted values
e_hat <- resid(reduced_model_cleaned_logged)
y_hat <- fitted(reduced_model_cleaned_logged)

# Plot residuals
par(mfrow=c(4,3))
plot(data_cleaned$BPSysAve ~ y_hat, ylab="True Systolic Blood Pressure", xlab="Fitted")
plot(e_hat ~ y_hat, ylab="Residuals", xlab="Fitted")
for(i in c(2,3,4,5,6,7,8,9,10,11,12)){
  plot(e_hat ~ data_cleaned[,i], ylab="Residuals", xlab=names(data_cleaned)[i])
}
plot(e_hat, ylab="Residuals", xlab="Observation number")
qqnorm(e_hat)
qqline(e_hat)
```

```{r}
final_model = reduced_model_cleaned_logged
vif(final_model, type='predictor')
```

```{r}
trainLogBP = log(train$BPSysAve)
testLogBP = log(test$BPSysAve)

model1 = lm(trainLogBP ~ Gender + Age + BMI + Poverty + Pulse + TotChol + PhysActiveDays + Race1 + PhysActive + TVHrsDay + Gender:Age + Gender:BMI + Age:TotChol, data = train)
model2 = lm(testLogBP ~ Gender + Age + BMI + Poverty + Pulse + TotChol + PhysActiveDays + Race1 + PhysActive + TVHrsDay + Gender:Age + Gender:BMI + Age:TotChol, data = test)
summary(model1)
summary(model2)
vif(model1, type = 'predictor')
vif(model2, type = 'predictor')
```
```
